Staged Compilation With Two-Level Type Theory

### 2022-02-22 paper/paper.tex
stuff

%%
%% end of the preamble, start of the body of the document source.
\hypersetup{draft}
\begin{document}

@@

\begin{itemize}
  \item In Section \ref{sec:tour-of-2ltt} we present an informal syntax of two-level
    type theory, a dependent type theory with staging features. We look at basic
    use cases involving inlining control, partial evaluation and fusion
    optimizations. We also describe feature variations, enabling applications in
    monomorphization and memory layout control.
  \item In Section \ref{sec:formal-2ltt}, we specify the formal notion of models and
    syntax for the variant of 2LTT used in this paper, largely following \cite{twolevel}.
  \item In Section \ref{sec:staging-algorithm}, we review the standard presheaf
    model of 2LTT \cite[Section~2.5.3]{twolevel}, which lies over the syntactic
    category of the object theory.  We show that evaluation in this model yields
    a staging algorithm for closed types and terms. We then prove stability for
    staging, which means that staging the output of staging has no action. Then,
    we extend staging to open types and terms which may depend on object-level
    contexts.
  \item
    In Section \ref{sec:soundness} we show soundness of staging, which means that
    the output of staging is convertible to the input.  Staging together with
    its stability and soundness can be viewed as a \emph{strong conservativity}
    theorem of 2LTT over the object theory. This means that the possible
    object-level constructions in 2LTT are in bijection with the constructions
    in the object theory, and staging witnesses that meta-level constructions
    can be always computed away. This strengthens the weak notion of
    conservativity shown in \cite{capriotti2017models} and \cite{twolevel}.
  \item
    In Section \ref{sec:intensional-analysis}, we discuss possible interpretations
    of intensional code analysis in the presheaf model.
  \item To our knowledge, this is the first account of staged compilation in the
    presence of full-blown dependent types, with universes and large
    elimination. In particular, we allow unrestricted staging for types, so that
    types can be computed by metaprograms at compile time.
\end{itemize}

@@

This setup is fairly similar to the well-known monomorphization models in the
C++ and Rust programming languages. The evident difference is that in 2LTT we
can use a full type theory at compile time, as opposed to languages which are
less expressive and less principled. Additionally, we will see in Section
\ref{sec:intensional-analysis} that this setup is compatible with an induction principle for
$\Ty_0$, so that we can analyze the structure of runtime types.

@@

We review the presheaf model described in \cite[Section~2.5.3]{twolevel}.  This
will directly yield a closed staging algorithm, by evaluation of 2LTT types and
terms in the model. First, we give a higher-level overview of this semantics.

TODO explanation here.
%% Here, contexts are presheaves over the syntactic category of
%% the object theory.

@@

\noindent The basic scheme for internalization is as follows:
\begin{alignat*}{3}
  & \Gamma : \hCon                  && \hspace{1em}\text{is internalized as}\hspace{1em} \Gamma : \whset_\omega\\
  & \sigma : \hSub\,\Gamma\,\Delta  && \hspace{1em}\text{is internalized as}\hspace{1em} \sigma : \Gamma \to \Delta\\
  & A : \hTy_{1,j}\,\Gamma           && \hspace{1em}\text{is internalized as}\hspace{1em} A      : \Gamma \to \whset_j \\
  & t : \hTm_{1,j}\,\Gamma\,A        && \hspace{1em}\text{is internalized as}\hspace{1em} t      : (\gamma : \Gamma) \to A\,\gamma
\end{alignat*}
\subsubsection{Object-theoretic syntax} The syntax of the object theory is clearly
fully stable under object-theoretic substitution, so we can internalize all of
its type and term formers. We internalize object-level types and terms as
$\Ty_{\mbbo j} : \whset_0$ and $\Tm_{\mbbo j} : \Ty_{\mbbo j} \to
\whset_0$. $\Ty_{\mbbo j}$ is closed under type formers. For instance, we have
\[ \Nat_j : \Ty_{\mbbo j}\hspace{1.5em}\zero_j : \Tm_{\mbbo j}\,\Nat_j\hspace{1.5em}
    \suc_j : \Tm_{\mbbo j}\,\Nat_j \to \Tm_{\mbbo j}\,\Nat_j \]
together with $\NatElim$, and likewise we have all other type formers.
\subsubsection{Internal $\ev$} We also use an internal view of
$\ev$ which maps 2LTT syntax to internal values; i.e.\ we compose $\ev$ with
internalization. Below, note that the input to $\ev$ is \emph{external}, so we
mark $\ev$ as being parametrized by external input.
\begin{alignat*}{3}
  &\ev\,(\Gamma &&: \Con)                      &&: \whset_\omega \\
  &\ev\,(\sigma &&: \Sub\,\Gamma\,\Delta)      &&: \ev\,\Gamma \to \ev\,\Delta \\
  &\ev\,(A      &&: \Ty_{1,j}\,\Gamma)          &&: \ev\,\Gamma \to \whset_j\\
  &\ev\,(t      &&: \Tm_{1,j}\,\Gamma)          &&: (\gamma : \ev\,\Gamma) \to \ev\,A\,\gamma\\
  &\ev\,(A      &&: \Ty_{0,j}\,\Gamma)          &&: \ev\,\Gamma \to \Ty_{\mbbo_j}\\
  &\ev\,(t      &&: \Tm_{0,j}\,\Gamma\,A)       &&: (\gamma : \ev\,\Gamma) \to \Tm_{\mbbo j}\,(\ev\,A\,\gamma)
\end{alignat*}
\subsubsection{Object-level fragment of 2LTT}
The purely object-level syntactic fragment of 2LTT can be internalized as
follows. We define externally the presheaf of object-level types as
$|\Ty_{0,j}|\,a := \Ty_{0,j}\,\emb{a}$, and the displayed presheaf of
object-level terms over $\Ty_{0,j}$ as $|\Tm_{0,j}|\,\{a\}\,A :=
\Tm_{0,j}\,\emb{a}\,A$. Hence, internally we have $\Ty_{0,j} : \whset_0$ and
$\Tm_{0,j} : \Ty_{0,j} \to \whset_0$. $\Ty_{0,j}$ is closed under all type formers,
analogously as we have seen for $\Ty_{\mbbo j}$.

@@

We define a family of functions $\re$ from the 2LTT syntax to objects in
$\hato$. This will be related to the evaluation morphism $\ev$ in the relational
interpretation. In short, $\re$ restricts 2LTT syntax so that it can only depend
on object-level contexts, i.e.\ contexts given as $\emb{\Gamma}$.

@@

\subsubsection{Object-level type formers}
Lastly, object-level type formers are straightforward. For types, we need
$\emb{\ev\,A\,\gamma} = \re\,A\,\gamma'$, and for terms, we need
$\emb{\ev\,A\,\gamma} = \re\,A\,\gamma'$. Note that $\ev$ and $\emb{\blank}$
preserve all structure, and $\re$ preserves all structure on object-level types
and terms. Hence, each object-level $A$ and $t$ case in $\blank\rel$ trivially
follows from induction hypotheses.

@@

If we have sufficiently expressive inductive types in the meta-level language,
it is possible to simply use inductive deeply embedded syntaxes, which can be
analyzed; our example in Section \ref{sec:staged-programming} for partially evaluating simply
typed lambda calculus is like this. However, native metaprogramming features
could be more concise and convenient than deep embeddings. Hence, intensional
analysis could provide additional convenience.

@@

Consider a very basic feature of intensional analysis, \emph{decidability of
definitional equality} for object-level Boolean expressions.
\[
  \msf{decEq} : (x\,y : \Lift\,\Bool_0) \to (x =_1 y) +_1 (x \neq_1 y)
\]
However, Yoneda lemma says that $\msf{decEq}$ cannot actually decide
definitional equality. We expect that $\msf{decEq}$ allows us to define
non-constant maps from $\Lift\,\Bool_0$ to $\Bool_1$, but the Yoneda lemma says
that we only have two terms with distinct semantics in $\Tm_1\,(\emptycon \ext_0
\Bool_0)\,\Bool_1$.

@@

However, this is still sufficient for many practical use cases, for example the
monomorphization setup in Section \ref{sec:monomorphization}. Here, it makes sense to only
having weakening in the object theory, but no $\beta\eta$-rules. We care about
the intension of the staging output; presumably we are doing monomorphization
because we care about code performance, so in 2LTT we do not want to
definitionally equate object-level programs with different performance
characteristics.

\subsubsection{Closed modalities}\label{sec:closed-modalities}
Another option is to add a \emph{closed} or \emph{crisp} modality
\cite{licata2018internal} for tracking object-level values which will become
closed terms after staging. Since closed terms are not affected by substitution,
we are free to analyze their intension. Such analysis can be viewed as an
internalization of canonicity for closed object-level terms. For example, a
closed $\Bool_0$ term will be either $\true_0$ or $\false_0$ after staging, so
it makes sense to be able to eliminate from it to $\Bool_1$. This can be also
viewed as a safe implementation of a \emph{run} function in the style of MetaML
\cite{metaml}, where \emph{scope extrusion} is prevented by the closed modality.

C-style \emph{function pointers} could be also an interesting use case, since
these have to become closed after staging (as they cannot capture closure
environments).

@@

The work Annekov et al.\ on two-level type theory \cite{twolevel}, building on
the thesis of Capriotti \cite{capriotti2017models} and ideas from Voevodsky
\cite{hts}, is the foremost inspiration in the current work. Interestingly, the
above works do mention metaprogramming as a source of intuition for 2LTT, but
they only briefly touch on this aspect, and in \cite{twolevel} the main focus is
on extensions of basic 2LTT which do not have staging semantics
anymore. Ibid.\ conjectures the strong conservativity of 2LTT in Proposition
2.18, and suggests categorical glueing as proof strategy. In this work, we do
essentially that: our presheaf model and the logical relation model could be
merged into a single instance of gluing along the $\re$ morphism; this merges
staging and its soundness proof. We split this construction to two parts in order
to get a self-contained presentation of the staging algorithm.

@@

The main difference is the explicit stage tracking via universes. Existing
systems all support type lifting (usually called $\msf{Code}$), quoting and
splicing, but lifting does not change the universe of types. For example, we can
write a function with type $\Bool \to \msf{Code}\,\Bool$, and the two $\Bool$
occurrences refer to the exact same type.

This yields some semantic and operational confusion down the line, and some
additional disambiguation is needed for stages of binders. For example, in typed
Template Haskell, top-level binders behave as runtime binders when used under a
quote, but behave as static binders otherwise. Given a top-level definition $b =
\True$, if we write a top-level definition $\msf{expr} = [||\,b\,||]$, then $b$
is a \emph{variable} in the staging output, but if we write $\msf{expr} =
\msf{if}\,b\,\msf{then}\,[||\,\True\,||]\,\msf{else}\,[||\,\False\,||]$, then
$b$ is considered to be a static value. In contrast, 2LTT does not distinguish
top-level and local binders, and in fact it has no syntactic or scope-based
restrictions; everything is enforced by typing.

In MetaML and MetaOCaml, there is a \emph{run} function for evaluating closed
object-level terms: as we mentioned in Section \ref{sec:closed-modalities}, we
believe that the proper way to implement it in 2LTT would be to use a closed
modality. This use case is closely related to \cite{licata2018internal}. In
ibid.\ a semantics is sketched for this modality in a presheaf topos, and we
could quite directly adapt it to our presheaf model of 2LTT. That said, working
out the practical details of this modality remains future work.

@@

Additionally, 2LTT only supports two stages, while the other systems all
allow countable stages. It remains future work to develop NLTT (N-level type
theory), but this extension does appear to be straightforward. In the
specification of NLTT, we would simply need to move from $i \in \{0,1\}$ to
$i \in \mbb{N}$, although in the semantics there could be more subtleties.

The multi-stage calculus of Kawata and Igarashi \cite{multi-stage-calculus}
supports staging operations and some type dependencies. However, there are major
differences to the current work. First, ibid.\ does not have universes or large
elimination, nor does it allow staged manipulation of type expressions. Second,
it does not support staging in the sense of Definition \ref{def:staging};
rather, it is a system for runtime code manipulation and evaluation. In other
words, programs may freely manipulate code fragments during execution, in more
liberal ways than in 2LTT, but there is no judgment which ensures that a purely
``object-level'' program can be extracted from a multi-stage program.

We emphasize that a distinction exists between staged compilation and staged
computation. For the former, we expect staging in the sense of Definition
\ref{def:staging}; this is essential if we want to use staging as a lowering
step in a compilation pipeline. Staged computation on the other hand may support
runtime code construction and evaluation, but may not guarantee that these
features can be computed away.

@@

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

>>>>

%%
%% end of the preamble, start of the body of the document source.
%% \hypersetup{draft}
\begin{document}

@@

\begin{itemize}
  \item In Section \ref{sec:tour-of-2ltt} we present an informal syntax of two-level
    type theory, a dependent type theory with staging features. We look at basic
    use cases involving inlining control, partial evaluation and fusion
    optimizations. We also describe feature variations, enabling applications in
    monomorphization and memory layout control.
  \item In Section \ref{sec:formal-2ltt}, we specify the formal notion of models and
    syntax for the variant of 2LTT used in this paper, largely following \cite{twolevel}.
  \item In Section \ref{sec:staging-algorithm}, we review the standard presheaf
    model of 2LTT \cite[Section~2.5.3]{twolevel}, which lies over the syntactic
    category of the object theory.  We show that evaluation in this model yields
    a staging algorithm for closed types and terms. We then prove stability for
    staging, which means that staging the output of staging has no action. Then,
    we extend staging to open types and terms, which may depend on object-level
    contexts.
  \item
    In Section \ref{sec:soundness} we show soundness of staging, which means that
    the output of staging is convertible to the input.  Staging together with
    its stability and soundness can be viewed as a \emph{strong conservativity}
    theorem of 2LTT over the object theory. This means that the possible
    object-level constructions in 2LTT are in bijection with the constructions
    in the object theory, and staging witnesses that meta-level constructions
    can be always computed away. This strengthens the weak notion of
    conservativity shown in \cite{capriotti2017models} and \cite{twolevel}.
  \item
    In Section \ref{sec:intensional-analysis}, we discuss possible interpretations
    of intensional code analysis in the presheaf model.
  \item We provide a small prototype implementation which we briefly discuss in
    Section \ref{sec:implementation}.
  \item To our knowledge, this is the first account of staged compilation in the
    presence of full-blown dependent types, with universes and large
    elimination. In particular, we allow unrestricted staging for types, so that
    types can be computed by metaprograms at compile time.
\end{itemize}

@@

This setup is fairly similar to the well-known monomorphization models in the
C++ and Rust programming languages. The evident difference is that in 2LTT we
can use a full type theory at compile time. Additionally, we will see in Section
\ref{sec:intensional-analysis} that this setup is compatible with an induction principle for
$\Ty_0$, so that we can analyze the structure of runtime types.

@@

We review the presheaf model described in \cite[Section~2.5.3]{twolevel}.  This
will directly yield a closed staging algorithm, by evaluation of 2LTT types and
terms in the model.

We give a higher-level overview first. Presheaves can be viewed as a
generalization of sets, as ``variable sets'' which may vary over a base
category. A set-based model of 2LTT yields a standard interpreter, where every
syntactic object is mapped to the obvious semantic counterpart. The presheaf
model over the category of contexts and substitutions in $\mbbo$ generalizes
this to an interpreter where semantic values may depend on object-theoretic
contexts. This allows us to handle object-theoretic \emph{types and terms}
during interpretation, since these necessarily depend on their contexts.

In the presheaf model, every semantic construction must be \emph{stable} under
object-level substitution, or using alternative terminology, must be
\emph{natural} with respect to substitution. Semantic values can be viewed as
runtime objects during interpretation, and naturality means that
object-theoretic substitutions can be \emph{applied} to such runtime objects,
acting on embedded object-theoretic types and terms. Stability captures the core
idea of staging:
\begin{itemize}
\item On one hand, every syntactic rule and construction in 2LTT must
      be stable, which restricts the range of metaprograms
      that can be written in 2LTT. For example, we cannot make decisions
      based on sizes of the object-theoretic scopes, since these sizes may be changed
      by substitution.
\item On the other hand, if only stable constructions are possible, we never
      have to \emph{prove} stability, or even explicitly mention object-theoretic
      contexts, substitutions and variables. The tedious details of deep embeddings
      can be replaced with a handful of staging operations and typing rules.
\end{itemize}
In principle, the base category in a 2LTT can be any category with finite
products; it is not required that it is a category of contexts and substitutions.
If the base category has more structure, we get more interesting object
theories, but at the same time stability becomes more onerous, because we must
be stable under \emph{more} morphisms. If the base category is simpler, with
fewer morphisms, then the requirement of stability is less restrictive. In
Section \ref{sec:intensional-analysis}, we will look at a concrete example for this.

@@

\noindent The basic scheme for internalization is as follows:
\begin{alignat*}{3}
  & \Gamma : \hCon                  && \hspace{1em}\text{is internalized as}\hspace{1em} \Gamma : \whset_\omega\\
  & \sigma : \hSub\,\Gamma\,\Delta  && \hspace{1em}\text{is internalized as}\hspace{1em} \sigma : \Gamma \to \Delta\\
  & A : \hTy_{1,j}\,\Gamma           && \hspace{1em}\text{is internalized as}\hspace{1em} A      : \Gamma \to \whset_j \\
  & t : \hTm_{1,j}\,\Gamma\,A        && \hspace{1em}\text{is internalized as}\hspace{1em} t      : (\gamma : \Gamma) \to A\,\gamma
\end{alignat*}
\subsubsection{Object-theoretic syntax} The syntax of the object theory is clearly
fully stable under object-theoretic substitution, so we can internalize all of
its type and term formers. We internalize object-level types and terms as
$\Ty_{\mbbo j} : \whset_0$ and $\Tm_{\mbbo j} : \Ty_{\mbbo j} \to
\whset_0$. $\Ty_{\mbbo j}$ is closed under type formers. For instance, we have
\[ \Nat_j : \Ty_{\mbbo j}\hspace{1.5em}\zero_j : \Tm_{\mbbo j}\,\Nat_j\hspace{1.5em}
    \suc_j : \Tm_{\mbbo j}\,\Nat_j \to \Tm_{\mbbo j}\,\Nat_j \]
together with $\NatElim$, and likewise we have all other type formers.
\subsubsection{Internal $\ev$} We also use an internal view of
$\ev$ which maps 2LTT syntax to internal values; i.e.\ we compose $\ev$ with
internalization. Below, note that the input to $\ev$ is \emph{external}, so we
mark $\ev$ as being parametrized by external input.
\begin{alignat*}{3}
  &\ev\,(\Gamma &&: \Con)                      &&: \whset_\omega \\
  &\ev\,(\sigma &&: \Sub\,\Gamma\,\Delta)      &&: \ev\,\Gamma \to \ev\,\Delta \\
  &\ev\,(A      &&: \Ty_{1,j}\,\Gamma)          &&: \ev\,\Gamma \to \whset_j\\
  &\ev\,(t      &&: \Tm_{1,j}\,\Gamma)          &&: (\gamma : \ev\,\Gamma) \to \ev\,A\,\gamma\\
  &\ev\,(A      &&: \Ty_{0,j}\,\Gamma)          &&: \ev\,\Gamma \to \Ty_{\mbbo_j}\\
  &\ev\,(t      &&: \Tm_{0,j}\,\Gamma\,A)       &&: (\gamma : \ev\,\Gamma) \to \Tm_{\mbbo j}\,(\ev\,A\,\gamma)
\end{alignat*}
\subsubsection{Object-level fragment of 2LTT}
The purely object-level syntactic fragment of 2LTT can be internalized as
follows. We define externally the presheaf of object-level types as
$|\Ty_{0,j}|\,\Gamma := \Ty_{0,j}\,\emb{\Gamma}$, and the displayed presheaf of
object-level terms over $\Ty_{0,j}$ as $|\Tm_{0,j}|\,\{\Gamma\}\,A :=
\Tm_{0,j}\,\emb{\Gamma}\,A$. Hence, internally we have $\Ty_{0,j} : \whset_0$ and
$\Tm_{0,j} : \Ty_{0,j} \to \whset_0$. $\Ty_{0,j}$ is closed under all type formers,
analogously as we have seen for $\Ty_{\mbbo j}$.

@@

We define a family of functions $\re$ from the 2LTT syntax to objects in
$\hato$. This will be related to the evaluation morphism $\ev$ in the relational
interpretation. In short, $\re$ restricts 2LTT syntax so that it can only depend
on object-level contexts, contexts given as $\emb{\Gamma}$.

@@

\subsubsection{Object-level type formers}
Lastly, object-level type formers are straightforward. For types, we need
$\emb{\ev\,A\,\gamma} = \re\,A\,\gamma'$, and likewise for terms. Note that
$\ev$ and $\emb{\blank}$ preserve all structure, and $\re$ preserves all
structure on object-level types and terms. Hence, each object-level $A$ and $t$
case in $\blank\rel$ trivially follows from induction hypotheses.

@@

If we have sufficiently expressive inductive types in the meta-level language,
it is possible to simply use inductive deeply embedded syntaxes, which can be
analyzed; our example in Section \ref{sec:staged-programming} for partially
evaluating simply typed lambda calculus is like this. However, native
metaprogramming features could be more concise and convenient than deep
embeddings, similarly to how staged programming is more convenient than code
generation with deeply embedded syntaxes. Hence, intensional analysis could
provide additional convenience.

@@

Consider a very basic feature of intensional analysis, \emph{decidability of
definitional equality} for object-level Boolean expressions.
\[
  \msf{decEq} : (x\,y : \Lift\,\Bool_0) \to (x =_1 y) +_1 (x \neq_1 y)
\]
However, the Yoneda lemma says that $\msf{decEq}$ cannot actually decide
definitional equality. We expect that $\msf{decEq}$ lets us define non-constant
maps from $\Lift\,\Bool_0$ to $\Bool_1$, but the Yoneda lemma implies that we
only have two terms with distinct semantics in $\Tm_1\,(\emptycon \ext_0
\Bool_0)\,\Bool_1$.

@@

However, this is still sufficient for many practical use cases, for example the
monomorphization setup in Section \ref{sec:monomorphization}. Here, it makes sense to only
have weakening in the object theory, but no $\beta\eta$-rules. We care about
the intension of the staging output; presumably we are doing monomorphization
because we care about code performance, so in 2LTT we do not want to
definitionally equate object-level programs with different performance
characteristics.

\subsubsection{Closed modalities}\label{sec:closed-modalities}
Another option is to add a \emph{closed} or \emph{crisp} modality
\cite{licata2018internal} for tracking object-level values which will become
closed terms after staging. Since closed terms are not affected by substitution,
we are free to analyze them. Such analysis can be viewed as an internalization
of canonicity for closed object-level terms. For example, a closed $\Bool_0$
term will be either $\true_0$ or $\false_0$ after staging, so it makes sense to
be able to eliminate from it to $\Bool_1$. This can be also viewed as a safe
implementation of a \emph{run} function in the style of MetaML \cite{metaml}.
In our case \emph{scope extrusion} would be prevented by the closed modality.

C-style \emph{function pointers} could be also an interesting use case, since
these must become closed after staging (as they cannot capture closure
environments).

@@

The work Annekov et al.\ on two-level type theory \cite{twolevel}, building on
the thesis of Capriotti \cite{capriotti2017models} and ideas from Voevodsky
\cite{hts}, is the foremost inspiration in the current work. Interestingly, the
above works do mention metaprogramming as a source of intuition for 2LTT, but
they only briefly touch on this aspect, and in \cite{twolevel} the main focus is
on extensions of basic 2LTT which do not have staging semantics
anymore. Ibid.\ conjectures the strong conservativity of 2LTT in Proposition
2.18, and suggests categorical gluing as proof strategy. In this work, we do
essentially that: our presheaf model and the logical relation model could be
merged into a single instance of gluing along the $\re$ morphism; this merges
staging and its soundness proof. We split this construction to two parts in order
to get a self-contained presentation of the staging algorithm.

The multi-stage calculus of Kawata and Igarashi \cite{multi-stage-calculus}
supports staging operations and type dependencies. However, there are major
differences to the current work. First, ibid.\ does not have universes or large
elimination, nor does it allow staged manipulation of type expressions. Second,
it does not support staging in the sense of Definition \ref{def:staging};
rather, it is a system for runtime code manipulation and evaluation. In other
words, programs may manipulate code fragments during execution, in more liberal
ways than in 2LTT, but there is no judgment which ensures that a purely
``object-level'' program can be extracted from a multi-stage program. We make a
disctinction between staged compilation and staged computation. For the former,
we expect staging in the sense of Definition \ref{def:staging}; this is
essential if we want to use staging as a lowering step in a compilation
pipeline. Staged computation on the other hand may support runtime code
construction and evaluation, but may not guarantee that these features can be
entirely computed away.

Idris 1 supports compile-time partial evaluation, through \texttt{static}
annotations on function arguments \cite{scrap-your-inefficient-engine}. This can
be used for dependently typed code generation, but Idris 1 does not guarantee
that partial evaluation adequately progresses. For example, we can pass a neutral
value as static argument in Idris 1, while in 2LTT static function arguments
are always canonical during staging.

@@

2LTT explicitly track stages of types, in contrast to the mentioned
systems. There, we have type lifting (usually called $\msf{Code}$), quoting and
splicing, but lifting does not change the universe of types. We can write a
function with type $\Bool \to \msf{Code}\,\Bool$, and the two $\Bool$
occurrences refer to the exact same type.

This yields some operational confusion down the line, and additional
disambiguation is needed for stages of binders. For example, in typed Template
Haskell, top-level binders behave as runtime binders when used under a quote,
but behave as static binders otherwise. Given a top-level definition $b =
\True$, if we write a top-level definition $\msf{expr} = [||\,b\,||]$, then $b$
is a \emph{variable} in the staging output, but if we write $\msf{expr} =
\msf{if}\,b\,\msf{then}\,[||\,\True\,||]\,\msf{else}\,[||\,\False\,||]$, then
$b$ is considered to be a static value. In contrast, 2LTT does not distinguish
top-level and local binders, and in fact it has no syntactic or scope-based
restrictions; everything is enforced by typing.

2LTT supports staging for types and dependent types, while to our knowledge no
previous system for staged compilation does the same. It appears to the author
that the main missing component in previous systems is \emph{dependent types at
the meta level}, rather than staging operations; staging operations in 2LTT and
MetaML are already quite similar. For example, any interesting operation on
the static-length vectors in Section \ref{sec:staged-programming} requires
meta-level dependent types, even if the object language is simply typed.

In MetaML and MetaOCaml, there is a \emph{run} function for evaluating closed
object-level terms: as we mentioned in Section \ref{sec:closed-modalities}, we
believe that the proper way to implement it in 2LTT would be to use a closed
modality. This use case is closely related to \cite{licata2018internal}. In
ibid.\ a semantics is sketched for this modality in a presheaf topos, and we
could directly adapt it to our presheaf model of 2LTT. That said, working
out the practical details of this modality remains future work.

Additionally, 2LTT only supports two stages, while the other noted systems allow
countable stages. It remains future work to develop NLTT (N-level type theory),
but this extension does appear to be straightforward. In the specification of
NLTT, we would simply need to move from $i \in \{0,1\}$ to $i \in \mbb{N}$,
although in the semantics there could be more subtleties.

@@

\section{Implementation}\label{sec:implementation}

@@

TODO: psh explain, intro overview upd, conclusion, implementation description, related work type staging

\bibliographystyle{ACM-Reference-Format}
\interlinepenalty=10000
\bibliography{references}
